---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Mark E. Drummond"
date: '2015-05-23'
output: 
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

# Overview

We analyze the response in guinea pig tooth growth to vitamin C supplementation. From the data set description:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

## Requirements

1. Load the ToothGrowth data and perform some basic exploratory data analyses 
2. Provide a basic summary of the data.
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions. 

# Exploratory Data Analysis

First we clean up our environment and set the seed for our random number generator to ensure consistent results during testing.

```{r}
# Clean up the working environment.
rm(list = ls())
# We set the seed for consistent results. Comment this line to allow for a
# random seed each time the code is run.
set.seed(19647)
```

Next we load the data and review the basic structure and summary statistics.

```{r}
data(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
head(ToothGrowth)
```

The dosage is actualy a factor so let's convert it to such:

```{r}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
str(ToothGrowth)
```

# Basic Summary

Each dosage / delivery method pair invovled 10 guinea pigs.

```{r}
table(ToothGrowth$dose, ToothGrowth$supp)
```

Comparing tooth growth against dosage:

```{r}
boxplot(ToothGrowth$len ~ ToothGrowth$dose,
        main = "Tooth Growth vs. Dosage",
        xlab = "Dosage",
        ylab = "Tooth Length")
```

Based on the plot, it seems reasonable to assumes a relationship between dosage and tooth growth.

Comparing tooth growth against vitamin C delivery method:

```{r}
boxplot(ToothGrowth$len ~ ToothGrowth$supp, 
        main = "Tooth Growth vs. Delivery Method",
        xlab = "Delivery Method",
        ylab = "Tooth Length")
```

The relationship here is less clear. While the bulk of the data related to orange juice as a delivery method is associated with increased tooth growth, the dispersion of the data for vitamin C tablets as a delivery method makes the situation less certain.

# Hypothesis Testing

```{r}
fit <- lm(len ~ dose + supp, data=ToothGrowth)
confint(fit)
```

# Conclusion